---
title: "cox_models"
author: "Daniel Johnson"
date: "2022-08-17"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_notebook:
    theme: journal
    toc: yes
    toc_depth: 2
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---
# Setup

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(coxme)
library(sjPlot)
library(sjlabelled)
library(sjmisc)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Data sources
```{r data_in}
load("C:/Users/dyjohnson/OneDrive/Documents/Research/SRP/full_list_clean.RData")
```

# Set up data
```{r setup_data}
full_list_cox <- full_list %>%
  filter(unique_date_stop <= final_removal_dt) %>%
  mutate(
    can_stat_active = factor(
      case_when(
        can_stat_active == 2110 ~ "Status 1",
        can_stat_active == 2120 ~ "Status 2",
        can_stat_active == 2130 ~ "Status 3",
        can_stat_active == 2140 ~ "Status 4",
        can_stat_active == 2150 ~ "Status 5",
        can_stat_active == 2160 ~ "Status 6"
      )
    ),
    can_stat_active = relevel(can_stat_active, ref = "Status 6"),
    rec_exception_status = factor(
      case_when(
        can_stat_active == "Status 1" &
          rec_exception == 0 ~ "Status 1 no exception",
        can_stat_active == "Status 1" &
          rec_exception == 1 ~ "Status 1 exception",
        can_stat_active == "Status 2" &
          rec_exception == 0 ~ "Status 2 no exception",
        can_stat_active == "Status 2" &
          rec_exception == 1 ~ "Status 2 exception",
        can_stat_active == "Status 3" &
          rec_exception == 0 ~ "Status 3 no exception",
        can_stat_active == "Status 3" &
          rec_exception == 1 ~ "Status 3 exception",
        can_stat_active == "Status 4" &
          rec_exception == 0 ~ "Status 4 no exception",
        can_stat_active == "Status 4" &
          rec_exception == 1 ~ "Status 4 exception",
        can_stat_active == "Status 5" ~ "Status 5",
        TRUE ~ "Status 6"
      ),
      levels = c(
        "Status 1 no exception",
        "Status 1 exception",
        "Status 2 no exception",
        "Status 2 exception",
        "Status 3 no exception",
        "Status 3 exception",
        "Status 4 no exception",
        "Status 4 exception",
        "Status 5",
        "Status 6"
      )
    ),
    rec_exception_status = relevel(rec_exception_status, ref = "Status 6"),
    lvad = factor(
      case_when(
        CAN_VAD_TY == 2 ~ "LVAD",
        TRUE ~ "No LVAD"
      ),
      levels = c("LVAD",
                 "No LVAD")
    ),
    lvad = relevel(lvad, ref = "No LVAD"),
    except_type = factor(
      case_when(
        CAN_VAD_TY == 2 & rec_exception == 1 ~ "except_vad",
        rec_exception == 1 ~ "except_no_vad",
        TRUE ~ "no_except"
      )
    ),
    except_type = relevel(except_type, ref = "no_except"),
  )
```

# All statuses and exceptions
``` {r coxph_stat_ex_long}
coxph_stat_ex_long <- coxph(Surv(t_start, t_stop, death) ~ rec_exception_status, data = full_list_cox)

summary(coxph_stat_ex_long)

plot_model(
  coxph_stat_ex_long,
  show.values = TRUE,
  value.offset = .3,
  axis.labels = c(
    "Status 5",
    "Status 4 exception",
    "Status 4 no exception",
    "Status 3 exception",
    "Status 3 no exception",
    "Status 2 exception",
    "Status 2 no exception",
    "Status 1 exception",
    "Status 1 no exception"
  ),
  title = "Proportional Hazards Model",
  axis.title = "Hazard Ratio"
)

ggsave("coxph_stat_ex_long.tiff", dpi = 1200)
```

# All statuses and exceptions and random center effects
``` {r coxme_stat_ex_long}
coxme_stat_ex_long <- coxme(Surv(t_start, t_stop, death) ~ rec_exception_status + (1 | CAN_LISTING_CTR_ID), data = full_list_cox)

summary(coxme_stat_ex_long)

plot_model(
  coxme_stat_ex_long,
  show.values = TRUE,
  value.offset = .3,
  axis.labels = c(
    "Status 5",
    "Status 4 exception",
    "Status 4 no exception",
    "Status 3 exception",
    "Status 3 no exception",
    "Status 2 exception",
    "Status 2 no exception",
    "Status 1 exception",
    "Status 1 no exception"
  ),
  title = "Mixed Effects Model with Center Random Effects",
  axis.title = "Hazard Ratio"
)

ggsave("coxme_stat_ex_long.tiff", dpi = 1200)

coxme_stat_ex_long2 <- coxme(Surv(t_start, t_stop, death) ~ can_stat_active + rec_exception + (1 | CAN_LISTING_CTR_ID), data = full_list_cox)

summary(coxme_stat_ex_long2)
exp(confint(coxme_stat_ex_long2))

plot_model(
  coxme_stat_ex_long2,
  show.values = TRUE,
  value.offset = .3,
  axis.labels = c(
    "Exception",
    "Status 5",
    "Status 4",
    "Status 3",
    "Status 2",
    "Status 1"
  ),
  title = "Mixed Effects Model with Center Random Effects",
  axis.title = "Hazard Ratio",
  grid.breaks = c(0.1, 1, 10, 100)
)

ggsave("coxme_stat_ex_long2.tiff", dpi = 1200, width = 8400, height = 6000)
```

# Each status and exceptions and random center effects
```{r coxme_stat_ex}
# Status 1
full_list_cox_stat1 <- full_list_cox %>%
  filter(can_stat_active == "Status 1")

coxme_stat1_ex <- coxme(Surv(t_start, t_stop, death) ~ rec_exception + (1 | CAN_LISTING_CTR_ID), data = full_list_cox_stat1)

summary(coxme_stat1_ex)
exp(confint(coxme_stat1_ex))

# Status 2
full_list_cox_stat2 <- full_list_cox %>%
  filter(can_stat_active == "Status 2")

coxme_stat2_ex <- coxme(Surv(t_start, t_stop, death) ~ rec_exception + (1 | CAN_LISTING_CTR_ID), data = full_list_cox_stat2)

summary(coxme_stat2_ex)
exp(confint(coxme_stat2_ex))

# Status 3
full_list_cox_stat3 <- full_list_cox %>%
  filter(can_stat_active == "Status 3")

coxme_stat3_ex <- coxme(Surv(t_start, t_stop, death) ~ rec_exception + (1 | CAN_LISTING_CTR_ID), data = full_list_cox_stat3)

summary(coxme_stat3_ex)
exp(confint(coxme_stat3_ex))

# Status 4
full_list_cox_stat4 <- full_list_cox %>%
  filter(can_stat_active == "Status 4")

coxme_stat4_ex <- coxme(Surv(t_start, t_stop, death) ~ rec_exception + (1 | CAN_LISTING_CTR_ID), data = full_list_cox_stat4)

summary(coxme_stat4_ex)
exp(confint(coxme_stat4_ex))

# Plot
coxme_stats_ex <- plot_models(
  coxme_stat1_ex,
  coxme_stat2_ex,
  coxme_stat3_ex,
  coxme_stat4_ex,
  axis.labels = "Exception",
  m.labels = c("Status 1", "Status 2", "Status 3", "Status 4"),
  # title = "Mixed Effects Model with Center Random Effects",
  legend.title = "",
  axis.title = "Hazard Ratio of Pre-Transplant Mortality",
  spacing = 1,
  show.values = TRUE,
  show.p = TRUE,
  grid.breaks = c(0.1, 1, 10)
)

coxme_stats_ex$layers[[4]]$aes_params$hjust <- 0.4

coxme_stats_ex 

ggsave("coxme_stats_ex2.tiff", dpi = 1200, width = 8400, height = 6000)
```

# All statuses and exceptions, LVAD interaction, and random center effects
``` {r coxme_lvad}
coxme_lvad <- coxme(Surv(t_start, t_stop, death) ~ can_stat_active + rec_exception*lvad + (1 | CAN_LISTING_CTR_ID), data = full_list_cox)

summary(coxme_lvad)

plot_model(
  coxme_lvad,
  show.values = TRUE,
  value.offset = .3,
  axis.labels = c(
    "Exception * LVAD",
    "LVAD",
    "Exception",
    "Status 5",
    "Status 4",
    "Status 3",
    "Status 2",
    "Status 1"
  ),
  title = "Mixed Effects Model with Center Random Effects and LVAD Interaction",
  axis.title = "Hazard Ratio"
)

ggsave("coxme_lvad.tiff", dpi = 1200, width = 8400, height = 6000)

coxme_lvad2 <- coxme(Surv(t_start, t_stop, death) ~ can_stat_active + except_type + (1 | CAN_LISTING_CTR_ID), data = full_list_cox)

summary(coxme_lvad2)
exp(confint(coxme_lvad2))

plot_model(
  coxme_lvad2,
  show.values = TRUE,
  value.offset = .3,
  axis.labels = c(
    "Exception with LVAD",
    "Exception without LVAD",
    "Status 5",
    "Status 4",
    "Status 3",
    "Status 2",
    "Status 1"
  ),
  title = "Mixed Effects Model with Center Random Effects and LVAD Interaction",
  axis.title = "Hazard Ratio"
)

ggsave("coxme_lvad2.tiff", dpi = 1200, width = 8400, height = 6000)

coxme_lvad3 <- coxme(Surv(t_start, t_stop, death) ~ can_stat_active + except_type + (1 | CAN_LISTING_CTR_ID), data = full_list_cox %>% mutate(except_type = relevel(except_type, ref = "except_no_vad")))

summary(coxme_lvad3)
```

# Post-transplant Kaplan-Meier curves
``` {r post_km}
library(survival)
library(ggsurvfit)

# Filter the data to patients who were transplanted before 2021 and all subsequent follow-up observations
full_list_post_tx <- full_list %>%
  group_by(PX_ID) %>%
  filter(REC_TX_DT == min(REC_TX_DT) &
           REC_TX_DT < as.Date("2021-01-01")) %>%
  ungroup() %>%
  filter(unique_date_stop >= REC_TX_DT) %>%
  group_by(PX_ID) %>%
  summarise(PX_ID = PX_ID,
            death_dt = max(death_dt),
            REC_TX_DT = REC_TX_DT,
            can_stat_active = first(can_stat_active),
            rec_exception = first(rec_exception),
            CAN_LISTING_CTR_ID = CAN_LISTING_CTR_ID) %>%
  unique() %>%
  ungroup()

# Add survival time
full_list_post_tx <- full_list_post_tx %>%
  mutate(
    survival_time = as.numeric(difftime(death_dt, REC_TX_DT, units = "days")),
    death_post_tx = if_else(!is.na(survival_time) & survival_time <= 365, 1, 0),
    survival_time = if_else(survival_time <= 365 & !is.na(survival_time),
                            survival_time,
                            400)
  )
     
# Relabel rec_exception
full_list_post_tx <- full_list_post_tx %>%
  mutate(rec_exception = if_else(rec_exception == 0, "No exception", "Exception"))

# Tabulate post-transplant outcomes
full_list_post_tx %>% count(can_stat_active, rec_exception, death_post_tx)

# Kaplan-Meier curves
## All statuses
fit_exception_all <- survfit(
  Surv(survival_time, death_post_tx) ~ rec_exception,
  data = full_list_post_tx
) 

### Exception survival rate
sum(
  full_list_post_tx$rec_exception == "Exception" &
    full_list_post_tx$death_post_tx == 0
) /
  sum(full_list_post_tx$rec_exception == "Exception")

### Standard survival rate
sum(
  full_list_post_tx$rec_exception == "No exception" &
    full_list_post_tx$death_post_tx == 0
) /
  sum(full_list_post_tx$rec_exception == "No exception")

### Kaplan-Meier curve
tiff("km_curve_all.tiff", res = 1200, width = 8400, height = 6000, units = "px", pointsize = 12)

fit_exception_all %>%
  ggsurvplot(
    ylim = c(0.75, 1),
    xlim = c(0, 365),
    xlab = "Time (days)",
    break.x.by = 60,
    ggtheme = theme_bw(),
    pval = TRUE,
    pval.size = 4,
    pval.coord = c(330, 1),
    conf.int = TRUE,
    legend.title = "",
    legend.labs = c(
      "Exception",
      "No exception"
    ),
    risk.table = TRUE
  )

dev.off()

## Status 1
fit_exception_1 <- survfit(
  Surv(survival_time, death_post_tx) ~ rec_exception,
  data = full_list_post_tx %>% filter(can_stat_active == 2110)
) 

### Exception survival rate
sum(
  full_list_post_tx$can_stat_active == 2110 &
    full_list_post_tx$rec_exception == "Exception" &
    full_list_post_tx$death_post_tx == 0
) /
  sum(
    full_list_post_tx$can_stat_active == 2110 &
      full_list_post_tx$rec_exception == "Exception"
  )

### Standard survival rate
sum(
  full_list_post_tx$can_stat_active == 2110 &
    full_list_post_tx$rec_exception == "No exception" &
    full_list_post_tx$death_post_tx == 0
) /
  sum(
    full_list_post_tx$can_stat_active == 2110 &
      full_list_post_tx$rec_exception == "No exception"
  )

### Kaplan-Meier curve
tiff("km_curve_1.tiff", res = 1200, width = 8400, height = 6000, units = "px", pointsize = 12)

fit_exception_1 %>%
  ggsurvplot(
    ylim = c(0.75, 1),
    xlim = c(0, 365),
    xlab = "Time (days)",
    break.x.by = 60,
    ggtheme = theme_bw(),
    pval = TRUE,
    pval.size = 4,
    pval.coord = c(330, 1),
    conf.int = TRUE,
    legend.title = "",
    legend.labs = c(
      "Exception",
      "No exception"
    ),
    risk.table = TRUE
  )

dev.off()

## Status 2
fit_exception_2 <- survfit(
  Surv(survival_time, death_post_tx) ~ rec_exception,
  data = full_list_post_tx %>% filter(can_stat_active == 2120)
) 

### Exception survival rate
sum(
  full_list_post_tx$can_stat_active == 2120 &
    full_list_post_tx$rec_exception == "Exception" &
    full_list_post_tx$death_post_tx == 0
) /
  sum(
    full_list_post_tx$can_stat_active == 2120 &
      full_list_post_tx$rec_exception == "Exception"
  )

### Standard survival rate
sum(
  full_list_post_tx$can_stat_active == 2120 &
    full_list_post_tx$rec_exception == "No exception" &
    full_list_post_tx$death_post_tx == 0
) /
  sum(
    full_list_post_tx$can_stat_active == 2120 &
      full_list_post_tx$rec_exception == "No exception"
  )

### Kaplan-Meier curve
tiff("km_curve_2.tiff", res = 1200, width = 8400, height = 6000, units = "px", pointsize = 12)

fit_exception_2 %>%
  ggsurvplot(
    ylim = c(0.75, 1),
    xlim = c(0, 365),
    xlab = "Time (days)",
    break.x.by = 60,
    ggtheme = theme_bw(),
    pval = TRUE,
    pval.size = 4,
    pval.coord = c(330, 1),
    conf.int = TRUE,
    legend.title = "",
    legend.labs = c(
      "Exception",
      "No exception"
    ),
    risk.table = TRUE
  )

dev.off()

## Status 3
fit_exception_3 <- survfit(
  Surv(survival_time, death_post_tx) ~ rec_exception,
  data = full_list_post_tx %>% filter(can_stat_active == 2130)
) 

### Exception survival rate
sum(
  full_list_post_tx$can_stat_active == 2130 &
    full_list_post_tx$rec_exception == "Exception" &
    full_list_post_tx$death_post_tx == 0
) /
  sum(
    full_list_post_tx$can_stat_active == 2130 &
      full_list_post_tx$rec_exception == "Exception"
  )

### Standard survival rate
sum(
  full_list_post_tx$can_stat_active == 2130 &
    full_list_post_tx$rec_exception == "No exception" &
    full_list_post_tx$death_post_tx == 0
) /
  sum(
    full_list_post_tx$can_stat_active == 2130 &
      full_list_post_tx$rec_exception == "No exception"
  )

### Kaplan-Meier curve
tiff("km_curve_3.tiff", res = 1200, width = 8400, height = 6000, units = "px", pointsize = 12)

fit_exception_3 %>%
  ggsurvplot(
    ylim = c(0.75, 1),
    xlim = c(0, 365),
    xlab = "Time (days)",
    break.x.by = 60,
    ggtheme = theme_bw(),
    pval = TRUE,
    pval.size = 4,
    pval.coord = c(330, 1),
    conf.int = TRUE,
    legend.title = "",
    legend.labs = c(
      "Exception",
      "No exception"
    ),
    risk.table = TRUE
  )

dev.off()

## Status 4
fit_exception_4 <- survfit(
  Surv(survival_time, death_post_tx) ~ rec_exception,
  data = full_list_post_tx %>% filter(can_stat_active == 2140)
) 

### Exception survival rate
sum(
  full_list_post_tx$can_stat_active == 2140 &
    full_list_post_tx$rec_exception == "Exception" &
    full_list_post_tx$death_post_tx == 0
) /
  sum(
    full_list_post_tx$can_stat_active == 2140 &
      full_list_post_tx$rec_exception == "Exception"
  )

### Standard survival rate
sum(
  full_list_post_tx$can_stat_active == 2140 &
    full_list_post_tx$rec_exception == "No exception" &
    full_list_post_tx$death_post_tx == 0
) /
  sum(
    full_list_post_tx$can_stat_active == 2140 &
      full_list_post_tx$rec_exception == "No exception"
  )

### Kaplan-Meier curve
tiff("km_curve_4.tiff", res = 1200, width = 8400, height = 6000, units = "px", pointsize = 12)

fit_exception_4 %>%
  ggsurvplot(
    ylim = c(0.75, 1),
    xlim = c(0, 365),
    xlab = "Time (days)",
    break.x.by = 60,
    ggtheme = theme_bw(),
    pval = TRUE,
    pval.size = 4,
    pval.coord = c(330, 1),
    conf.int = TRUE,
    legend.title = "",
    legend.labs = c(
      "Exception",
      "No exception"
    ),
    risk.table = TRUE
  )

dev.off()

970/(144+970) # Exception
1540/(168+1540) # No Exception

```

# Post-transplant survival Cox model
```{r post_cox}
# Cox model
full_list_post_tx_cox <- full_list_post_tx %>%
  mutate(
    can_stat_active = factor(
      case_when(
        can_stat_active == 2110 ~ "Status 1",
        can_stat_active == 2120 ~ "Status 2",
        can_stat_active == 2130 ~ "Status 3",
        can_stat_active == 2140 ~ "Status 4",
        can_stat_active == 2150 ~ "Status 5",
        can_stat_active == 2160 ~ "Status 6"
      )
    ),
    can_stat_active = relevel(can_stat_active, ref = "Status 6")
  )

coxme_stat_ex_post <- coxme(Surv(survival_time, death_post_tx) ~ can_stat_active + rec_exception + (1 | CAN_LISTING_CTR_ID), data = full_list_post_tx_cox)

summary(coxme_stat_ex_post)
exp(confint(coxme_stat_ex_post))

plot_model(
  coxme_stat_ex_post,
  show.values = TRUE,
  value.offset = .3,
  axis.labels = c(
    "Exception",
    "Status 5",
    "Status 4",
    "Status 3",
    "Status 2",
    "Status 1"
  ),
  title = "Mixed Effects Model with Center Random Effects",
  axis.title = "Hazard Ratio",
  grid.breaks = c(0.1, 1, 10)
)

ggsave("coxme_stat_ex_post.tiff", dpi = 1200, width = 8400, height = 6000)
```

# Predicted 90-day mortality by status and exception
```{r mortality_rates}
# Calculate baseline hazards
baseline_haz <- basehaz(coxph_stat_ex_long)

# Extract the model coefficients
coef_values <- coef(coxph_stat_ex_long)

# Calculate the linear predictor (LP) values for each status level
LP_values <- predict(coxph_stat_ex_long, newdata = data.frame(rec_exception_status = c("Status 1 no exception", "Status 1 exception", "Status 2 no exception", "Status 2 exception", "Status 3 no exception", "Status 3 exception", "Status 4 no exception", "Status 4 exception", "Status 5", "Status 6")), type = "lp")

# Calculate the predicted survival probabilities at 90 days for each status level
surv_prob_90days <- exp(-exp(LP_values)*baseline_haz[90,1])

# Print the predicted survival probabilities at 90 days for each status level
print(surv_prob_90days)

# Compare survival probabilities
## Status 1
prop.test(c(
  surv_prob_90days[1] * sum(full_list_cox$rec_exception_status == "Status 1 no exception"),
  surv_prob_90days[2] * sum(full_list_cox$rec_exception_status == "Status 1 exception")
),
c(
  sum(full_list_cox$rec_exception_status == "Status 1 no exception"),
  sum(full_list_cox$rec_exception_status == "Status 1 exception")
))

## Status 2
prop.test(c(
  surv_prob_90days[3] * sum(full_list_cox$rec_exception_status == "Status 2 no exception"),
  surv_prob_90days[4] * sum(full_list_cox$rec_exception_status == "Status 2 exception")
),
c(
  sum(full_list_cox$rec_exception_status == "Status 2 no exception"),
  sum(full_list_cox$rec_exception_status == "Status 2 exception")
))

## Status 3
prop.test(c(
  surv_prob_90days[5] * sum(full_list_cox$rec_exception_status == "Status 3 no exception"),
  surv_prob_90days[6] * sum(full_list_cox$rec_exception_status == "Status 3 exception")
),
c(
  sum(full_list_cox$rec_exception_status == "Status 3 no exception"),
  sum(full_list_cox$rec_exception_status == "Status 3 exception")
))

## Status 4
prop.test(c(
  surv_prob_90days[7] * sum(full_list_cox$rec_exception_status == "Status 4 no exception"),
  surv_prob_90days[8] * sum(full_list_cox$rec_exception_status == "Status 4 exception")
),
c(
  sum(full_list_cox$rec_exception_status == "Status 4 no exception"),
  sum(full_list_cox$rec_exception_status == "Status 4 exception")
))
```