---
title: "data_prep"
author: "Daniel Johnson"
date: "2022-08-17"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_notebook:
    theme: journal
    toc: yes
    toc_depth: 2
    toc_float: yes
---
# Setup

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(lubridate)
library(haven)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Data sources

## Read in Standard Scientific Registry of Transplant Recipients (SRTR) SAF files
```{r data_in}
cand_thor <- read_sas("C:/Users/dyjohnson/Box/SAF 2022 Q2/pubsaf2206/cand_thor.sas7bdat")
JustFormHR <- read_sas("C:/Users/dyjohnson/Box/SAF 2022 Q2/SupplementalData2206/ThoracicRegistration/JustFormHR.sas7bdat")
StatusJustEpisode <- read_sas("C:/Users/dyjohnson/Box/SAF 2022 Q2/SupplementalData2206/ThoracicRegistration/StatusJustEpisode.sas7bdat")

load("C:/Users/dyjohnson/Box/Heart Data Pipeline/RData files/full_list_pubsaf2206.RData")
```

## Fix transplant center
```{r CAN_LISTING_CTR_ID}
full_list <- full_list %>%
  select(-CAN_LISTING_CTR_ID) %>%
  left_join(cand_thor %>%
              select(PX_ID, CAN_LISTING_CTR_ID),
            by = "PX_ID")
```

## Add application variables from JustFormHR
```{r JustFormHR_vars}
full_list <- full_list  %>%
  left_join(JustFormHR %>%
              select(JustId, FormEffectiveDt, FormExpirationDt, ApplicationStatus),
            by = "JustId")
```

## Filter events needed
```{r filter_data}
full_list <- full_list %>%
  filter(
    grepl(
      "CAN_LISTING_DT|PERS_OPTN_DEATH_DT|PERS_SSA_DEATH_DT|PERS_RESTRICT_DEATH_DT|CAN_REM_DT|CANHX_BEGIN_DT|REC_TX_DT|ChangeDt",
      unique_event
    )
  )
```

## Extract listing date
``` {r CAN_LISTING_DT}
full_list <- full_list %>%
  mutate(
    CAN_LISTING_DT = case_when(grepl("CAN_LISTING_DT", unique_event) ~ unique_date)
  ) %>%
  group_by(PX_ID) %>%
  fill(CAN_LISTING_DT, .direction = "downup") %>%
  ungroup()
```

## Save point 1
```{r}
save1 <- full_list
```

## Apply inclusion criteria
```{r inclusion_criteria}
cand_thor_short <- cand_thor %>%
  select(
    "PX_ID",
    "PERS_ID",
    "CAN_AGE_IN_MONTHS_AT_LISTING",
    "CAN_LISTING_DT",
    "CAN_REM_DT",
    "WL_ORG",
    "CAN_LISTING_CTR_ID",
    "CAN_REM_CD"
  )

cand_thor_short <- cand_thor_short %>%
  group_by(PERS_ID) %>%
  filter(
    CAN_AGE_IN_MONTHS_AT_LISTING >= 216 &
      CAN_LISTING_DT >= as.Date("2018-10-18") &
      CAN_LISTING_DT <= as.Date("2021-12-01")
  ) %>%
  ungroup()
```

## Remove all candidates who were simultaneously listed at numerous sites for a heart.
``` {r multi_listed}
# Identify candidates with multiple simultaneous listings and keep the most recent listing only
multi_listed <- cand_thor_short %>%
  group_by(PERS_ID) %>%
  arrange(CAN_LISTING_DT) %>%
  filter(
    sum(WL_ORG == "HR") > 1 &
      as.numeric(last(CAN_LISTING_DT) - first(CAN_REM_DT), units = "days") < 0 # Overlapping listings
  ) %>%
  slice_tail(n = 1)

# Identify candidates without multiple simultaneous listings
single_listed <- cand_thor_short %>%
  filter(!PERS_ID %in% multi_listed$PERS_ID)
```

## Select candidates from full_list
``` {r select_candidates}
full_list <- full_list %>%
  filter(PX_ID %in% multi_listed$PX_ID | PX_ID %in% single_listed$PX_ID)
```

## Select variables
``` {r select_variables}
full_list <- full_list %>%
  select(
    # cand_thor variables
    PX_ID,
    CAN_AGE_IN_MONTHS_AT_LISTING,
    CAN_LISTING_CTR_ID,
    CAN_LISTING_DT,
    CAN_INIT_STAT,
    CAN_REM_CD,
    CANHX_STAT_CD,
    CAN_VAD_TY,
    WL_ORG,
    
    # JustFormHR variables
    JustId,
    ApplicationStatus,
    Exception,
    FormEffectiveDt, 
    FormExpirationDt,
    RequestedCandStatCd,
    
    # heart_data_pipeline variables
    unique_event,
    unique_date,
    t_start,
    t_stop
  )
```

## Save the filtered data
```{r save_filtered_data}
setwd("C:/Users/dyjohnson/OneDrive/Documents/Research/SRP")

save(full_list, file = "full_list_filtered.RData")
```

# Data management

## Load filtered data
```{r load_filtered_data}
load("C:/Users/dyjohnson/OneDrive/Documents/Research/SRP/full_list_filtered.RData")
```

## Add CAN_LAST_ACT_STAT_DT as an event
``` {r CAN_LAST_ACT_STAT_DT}
# Create a dataframe with CAN_LAST_ACT_STAT_DT
last_active <- cand_thor %>%
  filter(!is.na(CAN_LAST_ACT_STAT_DT) & PX_ID %in% full_list$PX_ID) %>%
  mutate(unique_date = floor_date(CAN_LAST_ACT_STAT_DT, unit = "days"), 
         unique_event = "CAN_LAST_ACT_STAT_DT") %>%
  select(PX_ID, unique_date, unique_event)

# Merge CAN_LAST_ACT_STAT_DT dataframe with full_list
full_list <- full_list %>%
  bind_rows(last_active) %>%
  arrange(PX_ID, unique_date)

rm(last_active)

# If CAN_LAST_ACT_STAT_DT is on the same date as another event, merge the rows
duplicates <- full_list %>% 
  group_by(PX_ID, unique_date) %>% 
  filter(n() > 1) %>%
  mutate(unique_event = str_c(unique_event, collapse = ",")) %>%
  summarise(
    across(
      c(everything()),
      ~last(na.omit(.))
    )
  ) %>%
  ungroup()

uniques <- full_list %>%
  group_by(PX_ID, unique_date) %>% 
  filter(n() <= 1) %>%
  ungroup()

full_list <- uniques %>%
  bind_rows(duplicates) %>%
  arrange(PX_ID, unique_date)

rm(uniques, duplicates)

# If CAN_LAST_ACT_STAT_DT is on a new date, fill in the data from the previous row
full_list <- full_list %>%
  group_by(PX_ID) %>%
  fill(everything(), .direction = "down") %>%
  ungroup()
```

## Add a copy of CANHX_STAT_CD where a candidate's most recent status is carried through temporarily inactive periods
```{r can_stat_active}
full_list <- full_list %>%
  mutate(can_stat_active = factor(case_when(
    !(CANHX_STAT_CD %in% c(1999, 2999)) ~ CANHX_STAT_CD
    ))) %>%
  group_by(PX_ID) %>%
  fill(can_stat_active, .direction = "down") %>%
  ungroup()
```

## Add a variable for a candidate's minimum status to identify permanently inactive candidates and candidates listed using the previous classification system
``` {r can_stat_min}
full_list <- full_list %>%
  group_by(PX_ID) %>%
  mutate(
    can_stat_min = min(CANHX_STAT_CD)
  ) %>%
  ungroup()
```

## Drop inactive rows for candidates who list as inactive and become active later
``` {r drop_inactive}
full_list <- full_list %>% 
  filter(!is.na(can_stat_active) | can_stat_min %in% c(1999, 2999))
```

## Add a time-dependent variable indicating whether a candidate is receiving a status exception
```{r rec_exception}
# Remove justification forms that aren't for approved exceptions
full_list <- full_list %>%
  mutate(
    across(
      c(JustId, FormEffectiveDt, FormExpirationDt, RequestedCandStatCd, ApplicationStatus, Exception),
      ~ case_when(Exception == 1 & ApplicationStatus %in% c(8, 10, 12, 13, 14, 16, 19, 27) ~ .)
    )
  )

# Fill application data from JustFormHR through exception period
full_list <- full_list %>% 
  group_by(PX_ID) %>%
  fill(JustId, FormEffectiveDt, FormExpirationDt, RequestedCandStatCd, Exception, ApplicationStatus, .direction = "down") %>%
  mutate(
    across(
      c(JustId, FormEffectiveDt, FormExpirationDt, RequestedCandStatCd, Exception, ApplicationStatus),
      ~ case_when(unique_date <= FormExpirationDt ~ .)
    )
  ) %>%
  ungroup()

# status_approved represents a candidate's approved status and fills forward
full_list <- full_list %>%
  mutate(status_approved = case_when(
    Exception == 1 &
      ApplicationStatus %in% c(8, 10, 12, 13, 14, 16, 19, 27) ~ RequestedCandStatCd
    )
  ) %>%
  group_by(PX_ID) %>%
  fill(status_approved, .direction = "down") %>%
  ungroup()

full_list <- full_list %>%
  mutate(
    rec_exception = if_else(can_stat_active == status_approved & unique_date <= FormExpirationDt & unique_date + days(7) >= FormEffectiveDt, 1, 0, missing = 0)
  )
```

## Add a variable for whether a candidate ever received a status exception
``` {r rec_exception_ever}
full_list <- full_list %>%
  mutate(rec_exception_ever = case_when(rec_exception == 1 ~ 1)) %>%
  group_by(PX_ID) %>%
  fill(rec_exception_ever, .direction = "downup") %>%
  ungroup() %>%
  mutate(rec_exception_ever = tidyr::replace_na(rec_exception_ever, 0))
```

## Recalculate t_start and t_stop
```{r t_start_t_stop}
full_list <- full_list %>%
  ungroup() %>% 
  mutate(t_start = 0, t_stop = 0) %>% 
  arrange(PX_ID, unique_date)

for (i in 2:nrow(full_list)) {
  if (full_list$PX_ID[[i]] == 
      full_list$PX_ID[[i-1]]) {
    
    full_list$t_stop[[i-1]] = 
      full_list$unique_date[[i]] - 
      full_list$unique_date[[i-1]] +
      full_list$t_start[[i-1]]
    
    full_list$t_start[[i]] = full_list$t_stop[[i-1]]
    
  } else {
    
    full_list$t_stop[[i-1]] = 
      full_list$t_start[[i-1]]
    
    full_list$t_start[[i]] = 0
      
  }
}

full_list <- full_list %>% 
  relocate(t_start, t_stop, .after = unique_date)
```

## Save point 2
```{r}
save2 <- full_list
```

## Add length of statuses
```{r status_length}
full_list <- full_list %>%
  group_by(PX_ID) %>%
  mutate(
    can_stat_active_prev = lag(can_stat_active),
    can_stat_active_next = lead(can_stat_active),
    stat_start = case_when(
      can_stat_active != can_stat_active_prev | is.na(can_stat_active_prev) ~ t_start
    ),
    stat_stop = case_when(
      can_stat_active != can_stat_active_next | is.na(can_stat_active_next) ~ t_stop
    )
  ) %>%
  fill(stat_start, .direction = "down") %>%
  fill(stat_stop, .direction = "up") %>%
  mutate(stat_length = stat_stop - stat_start) %>% 
  select(-c(can_stat_active_prev, can_stat_active_next))
```

## Add length of exceptions
```{r exception_length}
full_list <- full_list %>%
  group_by(PX_ID) %>%
  mutate(
    rec_exception_prev = lag(rec_exception),
    rec_exception_next = lead(rec_exception),
    exception_start = case_when(
      rec_exception != rec_exception_prev | 
        is.na(rec_exception_prev) ~ t_start
    ),
    exception_stop = case_when(
      rec_exception != rec_exception_next |
        is.na(rec_exception_next) ~ t_stop
    )
  ) %>%
  fill(exception_start, .direction = "down") %>%
  fill(exception_stop, .direction = "up") %>%
  mutate(
    exception_start = case_when(rec_exception == 1 ~ exception_start),
    exception_stop = case_when(rec_exception == 1 ~ exception_stop),
    exception_length = case_when(rec_exception == 1 ~ exception_stop - exception_start) 
  ) %>% 
  select(-c(rec_exception_prev, rec_exception_next))
```

## Add a variable for whether a candidate received an exception at listing
``` {r rec_exception_init}
full_list <- full_list %>%
  mutate(rec_exception_init = case_when(rec_exception == 1 &
                                          t_start == 0 ~ 1)) %>%
  group_by(PX_ID) %>%
  fill(rec_exception_init, .direction = "downup") %>%
  ungroup() %>%
  mutate(rec_exception_init = tidyr::replace_na(rec_exception_init, 0))
```

## Add a variable for whether a candidate received an exception at listing or later
```{r rec_exception_init_ever}
full_list <- full_list %>%
  mutate(
    rec_exception_init_ever = factor(
      case_when(
        rec_exception_init == 1 ~ "Exception at listing",
        rec_exception_ever == 1 ~ "Exception after listing",
        TRUE ~ "No exception"
      ),
      levels = c("Exception at listing",
                 "Exception after listing",
                 "No exception")
    )
  )
```

## Add death date, which basically combines both the OPTN and SSA death dates
```{r death_dt}
full_list <- full_list %>%
  mutate(death_dt = case_when(
    grepl("PERS_OPTN_DEATH_DT", unique_event) |
      grepl("PERS_SSA_DEATH_DT", unique_event) |
      (grepl("CAN_REM_DT", unique_event) &
         CAN_REM_CD == 8) ~ unique_date
  )) %>%
  group_by(PX_ID) %>%
  fill(death_dt, .direction = "downup") %>%
  ungroup() 
```

## Add death
```{r death}
# Several candidates may be removed from the waitlist without getting a transplant but still die. This variable death accounts for that. Basically, if a candidate has a removal code for anything other than transplantation and has a death_dt on file, then they are coded as having died.
full_list <- full_list %>%
  mutate(death = case_when(
    unique_date >= death_dt &
      CAN_REM_CD %in% c(5, 6, 7, 8, 9, 10, 12, 13, 16, 17, 20, 24) ~ 1,
    TRUE ~ 0
  ))
```

## Add removal cause. If a candidate is not removed from the waitlist yet, then they are coded as 0. Death = 1. Transplant = 2. All other causes = 3.
```{r removal_cause}
full_list <- full_list %>%
  mutate(
    removal_cause = case_when(
      is.na(CAN_REM_CD) ~ 0,
      death == 1 ~ 1,
      CAN_REM_CD %in% c(4, 14, 15, 18, 19, 21, 22, 23) ~ 2,
      TRUE ~ 3
    )
  )
```

## Add removal date to account for censoring, death, and transplant
```{r final_removal_dt}
# Extract removal, and transplant dates
full_list <- full_list %>%
  mutate(
    CAN_REM_DT = case_when(grepl("CAN_REM_DT", unique_event) ~ unique_date),
    REC_TX_DT = case_when(grepl("REC_TX_DT", unique_event) ~ unique_date)
  ) %>%
  group_by(PX_ID) %>%
  fill(CAN_REM_DT, REC_TX_DT, .direction = "downup") %>%
  ungroup()

# If they were transplanted but don't have a transplant date on file, transplant date = removal date
full_list <- full_list %>%
  mutate(REC_TX_DT = coalesce(REC_TX_DT, CAN_REM_DT))

full_list <- full_list %>%
  group_by(PX_ID) %>%
  mutate(
    final_removal_dt = min(case_when(
      removal_cause == 0 ~ max(unique_date), # If they are still on the waitlist, final_removal_dt = most recent record date
      removal_cause == 1 ~ death_dt, # If they died, final_removal_dt = death date
      removal_cause == 2 ~ REC_TX_DT, # If they were transplanted, final_removal_dt = transplant date
      # removal_cause == 2 & is.na(REC_TX_DT) ~ CAN_REM_DT, # If they were transplanted but don't have a transplant date on file, final_removal_dt = removal date
      removal_cause == 3 ~ CAN_REM_DT # If they were removed for any other reason, final_removal_dt = removal date
    ),
    na.rm = TRUE)
  ) %>%
  ungroup()
```

## Delete deaths over 60 days after delisting
```{r trim_events}
full_list <- full_list %>%
  filter(
    unique_date <= final_removal_dt + days(60) # All events over 60 days after delisting are deaths
  )
```

## Add variable to indicate if a candidate is on the waitlist
``` {r on_waitlist}
full_list <- full_list %>%
  mutate(on_waitlist = if_else(unique_date < final_removal_dt, 1, 0))
```

## Add pre-transplant survival time
```{r survival_time}
full_list <- full_list %>%
  mutate(survival_time = as.numeric(difftime(final_removal_dt, CAN_LISTING_DT, units = "days")))
```

# Apply exclusion criteria
``` {r exclusions}
# Save data for CONSORT diagram before applying exclusion criteria
setwd("C:/Users/dyjohnson/OneDrive/Documents/Research/SRP")

save(full_list, file = "full_list_consort.RData")

# Exclusion criteria
full_list <- full_list %>%
  group_by(PX_ID) %>%
  filter(
    can_stat_min %in% c(2110, 2120, 2130, 2140, 2150, 2160, 2999) &
      can_stat_min != 2999 &
      WL_ORG == "HR"
  ) %>%
  ungroup()

# Add number of candidates per listing center
full_list <- full_list %>%
  group_by(CAN_LISTING_CTR_ID) %>%
  mutate(can_listing_ctr_count = n_distinct(PX_ID)) %>%
  ungroup()

# Filter out centers with fewer than 10 candidates
full_list <- full_list %>%
  filter(can_listing_ctr_count >= 10)
```

# Save data for the funnel plot before merging last two rows for each candidate
```{r save_funnel_data}
full_list_for_funnel <- full_list

setwd("C:/Users/dyjohnson/OneDrive/Documents/Research/SRP")

save(full_list_for_funnel, file = "full_list_for_funnel.RData")
```

# Combine the last two rows for each candidate so their final row isn't a single day
```{r final_row}
# Turn unique_date into unique_date_start and add unique_date_stop
full_list <- full_list %>%
  rename(unique_date_start = unique_date) %>% 
  arrange(PX_ID, unique_date_start) %>% 
  group_by(PX_ID) %>% 
  mutate(unique_date_stop = lead(unique_date_start)) %>%
  relocate(unique_date_stop, .after = unique_date_start) %>%
  ungroup()

# Merge the data in the last two rows for each candidate
full_list_head <- full_list %>%
  group_by(PX_ID) %>%
  slice_head(n = -2) %>%
  ungroup()

full_list_tail <- full_list %>%
  group_by(PX_ID) %>%
  slice_tail(n = 2) %>%
  mutate(unique_event = str_c(unique_event, collapse = ",")) %>%
  summarise(
    across(
      c(everything(),-unique_date_start,-t_start,-on_waitlist),
      ~last(na.omit(.))
    ),
    unique_date_start = min(unique_date_start, na.rm = TRUE),
    t_start = min(t_start, na.rm = TRUE),
    on_waitlist = first(on_waitlist)
  ) %>%
  ungroup() %>%
  relocate(unique_event, unique_date_start, .before = unique_date_stop) %>%
  relocate(t_start, .before = t_stop)

full_list <- full_list_head %>%
  bind_rows(full_list_tail) %>%
  arrange(PX_ID, unique_date_start)

rm(full_list_head, full_list_tail)
```

# Aggregate rows without changes in status, exception, or active status on waitlist
``` {r aggregate}
full_list <- full_list %>%
  group_by(PX_ID, rec_exception, can_stat_active, exception_start, stat_start, on_waitlist) %>%
  mutate(unique_event = str_c(unique_event, collapse = ",")) %>%
  summarise(
    across(
      c(everything(),-unique_date_start,-t_start),
      ~last(na.omit(.))
    ),
    unique_date_start = min(unique_date_start, na.rm = TRUE),
    t_start = min(t_start, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  relocate(unique_event, unique_date_start, .before = unique_date_stop) %>%
  relocate(t_start, .before = t_stop) %>% 
  relocate(exception_start, .before = exception_stop) %>%
  relocate(stat_start, .before = stat_stop) %>%
  arrange(PX_ID, unique_date_start) %>% 
  arrange(PX_ID, unique_date_start)
```

# Save final data
``` {r save_clean_data}
setwd("C:/Users/dyjohnson/OneDrive/Documents/Research/SRP")

save(full_list, file = "full_list_clean.RData")

write_csv(full_list, "full_list_clean.csv")
```

# Create sample data of 100 random candidates for supplement
``` {r sample_data}
full_list_sample_px <- full_list %>% select(PX_ID) %>% unique() %>% sample_n(100)

full_list_sample <- full_list %>% 
  filter(PX_ID %in% full_list_sample_px$PX_ID) %>%
  relocate(can_stat_active, .after = t_stop) %>%
  relocate(rec_exception, .after = status_approved) %>%
  relocate(on_waitlist, .after = final_removal_dt) %>%
  relocate(CAN_REM_DT, .after = CAN_REM_CD) %>%
  relocate(REC_TX_DT, .after = CANHX_STAT_CD)

setwd("C:/Users/dyjohnson/OneDrive/Documents/Research/SRP")

write_csv(full_list, "full_list_sample.csv")
```