---
title: "descriptive_plots"
author: "Daniel Johnson"
date: "2022-08-11"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_notebook:
    theme: journal
    toc: yes
    toc_depth: 2
    toc_float: yes
---
# Setup

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(ggthemes)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Data sources
```{r data_in}
load("C:/Users/dyjohnson/OneDrive/Documents/Research/SRP/full_list_clean.RData")
```

# Descriptive plots

## Exception frequency by status
``` {r exceptions_by_status}
## Create status_approved2, which is the same as status_approved but doesn't fill forward
full_list <- full_list %>%
  mutate(status_approved2 = case_when(
    Exception == 1 &
      ApplicationStatus %in% c(8, 10, 12, 13, 14, 16, 19, 27) &
      RequestedCandStatCd == 2110 ~ "Status 1",
    Exception == 1 &
      ApplicationStatus %in% c(8, 10, 12, 13, 14, 16, 19, 27) &
      RequestedCandStatCd == 2120 ~ "Status 2",
    Exception == 1 &
      ApplicationStatus %in% c(8, 10, 12, 13, 14, 16, 19, 27) &
      RequestedCandStatCd == 2130 ~ "Status 3",
    Exception == 1 &
      ApplicationStatus %in% c(8, 10, 12, 13, 14, 16, 19, 27) &
      RequestedCandStatCd == 2140 ~ "Status 4",
    Exception == 1 &
      ApplicationStatus %in% c(8, 10, 12, 13, 14, 16, 19, 27) &
      RequestedCandStatCd == 2150 ~ "Status 5",
    Exception == 1 &
      ApplicationStatus %in% c(8, 10, 12, 13, 14, 16, 19, 27) &
      RequestedCandStatCd == 2160 ~ "Status 6",
    ),
    status_approved2 = factor(status_approved2, levels = c(
          "Status 1",
          "Status 2",
          "Status 3",
          "Status 4",
          "Status 5",
          "Status 6"))
  )

d <- full_list %>%
  filter(
    !is.na(status_approved2) &
      status_approved2 %in% c("Status 1", "Status 2", "Status 3", "Status 4")
  )

# Counts
ggplot(d, aes(status_approved2)) + 
  geom_bar() +
  labs(x = "Status Exception Approved", y = "Count")

ggsave("status_approved.png", width = 5, height = 5)
```

## Exception proportion by status
``` {r status_approved_prop}
status_approved_prop <- full_list %>%
  filter(can_stat_active %in% c(2110, 2120, 2130, 2140)) %>%
  select(PX_ID,
         t_start,
         t_stop,
         can_stat_active,
         rec_exception) %>%
  mutate(
    can_stat_active = case_when(
      can_stat_active == 2110 ~ "Status 1",
      can_stat_active == 2120 ~ "Status 2",
      can_stat_active == 2130 ~ "Status 3",
      can_stat_active == 2140 ~ "Status 4",
    )
  ) %>%
  group_by(can_stat_active) %>%
  summarise(proportion_ex = mean(rec_exception) * 100) %>%
  ungroup()

ggplot(status_approved_prop, aes(can_stat_active, proportion_ex)) + 
  geom_col() +
  labs(x = "Status Exception Approved", y = "Proportion of Candidates with Exceptions")

ggsave("status_approved_prop.png", width = 5, height = 5)
```

## Stacked bar chart
``` {r status_stack}
## Add a time-dependent variable for whether an exception was received at listing or later
for_stack <- full_list %>%
  mutate(rec_exception_time = factor(
    case_when(
      rec_exception_init == 1 &
        t_start == 0 ~ "Exception at listing",
      rec_exception == 1 &
        t_start > 1 ~ "Exception after listing",
      TRUE ~ "No exception"
    ),
    levels = c("Exception at listing",
               "Exception after listing",
               "No exception")
  ))

for_stack <- for_stack %>%
  group_by(PX_ID) %>%
  mutate(
    rec_exception_status = factor(
      case_when(
        can_stat_active == 2110 &
          rec_exception == 0 ~ "Status 1 no exception",
        can_stat_active == 2110 &
          rec_exception == 1 ~ "Status 1 exception",
        can_stat_active == 2120 &
          rec_exception == 0 ~ "Status 2 no exception",
        can_stat_active == 2120 &
          rec_exception == 1 ~ "Status 2 exception",
        can_stat_active == 2130 &
          rec_exception == 0 ~ "Status 3 no exception",
        can_stat_active == 2130 &
          rec_exception == 1 ~ "Status 3 exception",
        can_stat_active == 2140 &
          rec_exception == 0 ~ "Status 4 no exception",
        can_stat_active == 2140 &
          rec_exception == 1 ~ "Status 4 exception",
        can_stat_active == 2150 ~ "Status 5",
        can_stat_active == 2160 ~ "Status 6"
      )
    ),
    can_stat_active = factor(
      case_when(
        can_stat_active == 2110 ~ "Status 1",
        can_stat_active == 2120 ~ "Status 2",
        can_stat_active == 2130 ~ "Status 3",
        can_stat_active == 2140 ~ "Status 4",
        can_stat_active == 2150 ~ "Status 5",
        can_stat_active == 2160 ~ "Status 6"
      ),
      levels = c(
        "Status 1",
        "Status 2",
        "Status 3",
        "Status 4",
        "Status 5",
        "Status 6"
      )
    ),
    rec_exception = factor(
      case_when(
        rec_exception == 1 ~ "Exception",
        TRUE ~ "No exception"),
      levels = c("Exception", "No exception")),
  ) %>%
  ungroup()
  
ggplot(for_stack, aes(x = can_stat_active, fill = rec_exception)) + 
  geom_bar() + 
  labs(x = "Status", y = "Number of Status Justifications", fill = "")

ggsave("status_stack.png", width = 8, height = 5)

plot <- ggplot(for_stack, aes(x = can_stat_active, fill = rec_exception_time)) + 
  geom_bar() + 
  labs(x = "Status", y = "Number of Status Justifications", fill = "")

plot +
  theme_gdocs() + 
  scale_fill_gdocs()

ggsave("status_stack_split.png", width = 8, height = 5)

for_stack %>% group_by(can_stat_active) %>% count(rec_exception_time)

table(for_stack$rec_exception_time)
```

# Chart of exceptions by time of exception
``` {r exception_by_time}
temp <- full_list %>%
  select(PX_ID,
         rec_exception,
         can_stat_active,
         status_approved,
         exception_start,
         exception_stop,
         exception_length) %>%
  filter(rec_exception == 1 &
           can_stat_active %in% c(2110, 2120, 2130, 2140)) %>%
  mutate(
    can_stat_active = case_when(
      can_stat_active == 2110 ~ "Status 1",
      can_stat_active == 2120 ~ "Status 2",
      can_stat_active == 2130 ~ "Status 3",
      can_stat_active == 2140 ~ "Status 4"
    )
  ) %>%
  distinct()

ggplot(temp, aes(can_stat_active, exception_start)) +
  geom_boxplot(outlier.shape = NA) +
  coord_cartesian(ylim = quantile(d$t_start, c(0.1, 0.9))) +
  labs(x = "Status Exception Approved", y = "Day Exception Approved after Listing")

ggplot(d, aes(status_approved2, t_start)) +
  geom_boxplot(outlier.shape = NA) +
  coord_cartesian(ylim = quantile(d$t_start, c(0.1, 0.9))) +
  labs(x = "Status Exception Approved", y = "Day Exception Approved")

ggsave("exception_by_time.png", width = 5, height = 5)
```

# Bar chart of exceptions by duration of exceptions
``` {r exception_by_length}
e <- full_list %>%
  filter(
    rec_exception == 1
  ) %>%
  select(PX_ID, exception_start, exception_stop, exception_length) %>%
  mutate(exception_length_mo = exception_length/30.437) %>%
  unique()

ggplot(e, aes(exception_length_mo)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Status Exception Length (months)", y = "Count")

ggsave("exception_by_length.png")
```

# Percentage of exception requests approved by region
``` {r exception_approval}
library(tidyverse)
library(haven)
library(geodist)
library(purrr)
library(ggmap)
library(tidycensus)
library(tigris)
library(sf)
library(usa)
library(ggspatial)
library(beepr)
library(censusxy)
library(tidygeocoder)
library(geosphere)
library(ggalt)
library(ggsci)
library(ggthemes)
library(ggrepel)
library(ggpubr)
library(RColorBrewer)
library(kableExtra)
library(weights)
library(gridExtra)
library(ggpubr)
library(grid)
```

```{r data cleaning, include = FALSE}
# # Add CAN_LISTING_CTR_ID to JustFormHR through JustId
JustFormHR <- read_sas("C:/Users/dyjohnson/Box/SAF 2022 Q2/SupplementalData2206/ThoracicRegistration/JustFormHR.sas7bdat")
cand_thor <- read_sas("C:/Users/dyjohnson/Box/SAF 2022 Q2/pubsaf2206/cand_thor.sas7bdat")

load("C:/Users/dyjohnson/Box/Heart Data Pipeline/RData files/full_list_pubsaf2206.RData")
```


``` {r}
full_list_ctr_fixed <- full_list %>%
  select(-CAN_LISTING_CTR_ID) %>%
  left_join(cand_thor %>%
              select(PX_ID, CAN_LISTING_CTR_ID),
            by = "PX_ID")

JustId_ctr_id <- full_list_ctr_fixed %>% filter(!is.na(JustId)) %>% select(JustId, CAN_LISTING_CTR_ID)

JustFormHR_ctr <- left_join(JustFormHR, JustId_ctr_id, by = "JustId") %>% unique()

exception_requests <- JustFormHR_ctr %>%
  filter(Exception == 1 & ApplicationStatus != 26) %>%
  mutate(approved = if_else(ApplicationStatus %in% c(8, 10, 12, 13, 14, 16, 18, 19, 27), 1, 0))

# centers with geocodes
center_long_lat <- read_csv("center_geocodes.csv") %>%
  rename(long = lon, CTR_NAME = ENTIRE_NAME) %>%
  mutate(CAN_LISTING_CTR_ID = CTR_ID)

# clean up NA and incorrectly coded regions

institution <- haven::read_sas("institution.sas7bdat")

opo2region <- institution %>%
  select(CTR_CD, CTR_ID, REGION)

load("opo_shapes.R") 

opo_shapes <- opo_shapes %>%
  rename(CTR_CD = OPO)

region_shapes <- inner_join(opo_shapes, opo2region) %>%
  group_by(REGION) %>% summarise(estimate = sum(estimate), do_union = TRUE)

df <- exception_requests %>%
  left_join(center_long_lat) %>%
  left_join(opo2region)

df_by_region <- df %>% 
  group_by(REGION) %>%
  summarise(region_requests = n(), region_approved = sum(approved)) %>%
  mutate(region_proportion = region_approved/region_requests)
  
chisq.test(df_by_region$region_proportion)
```

# Proportion of Exceptions at Time of Listing by Region

```{r region map}
regionmapproportion <- ggplot(df_by_region %>%
                                left_join(region_shapes)) +
  geom_sf(aes(fill = region_proportion, geometry = geometry), size = 0.2) +
  scale_fill_viridis_c(limits = c(0.93, 1)) +
  theme_void() +
  labs(fill = "Proportion of Exception\nRequests Approved") +
  theme(title = element_text(size = 10))

regionmapproportion

ggsave("regionmapproportion.png", regionmapproportion)
```


# Exception rates over time
``` {r exception_rates}
library(zoo)

# Convert the exception_date column to a Date object
exception_requests$FormEffectiveDt <- as.Date(exception_requests$FormEffectiveDt)

# Create a time series object with the number of exceptions per day
exceptions_ts <- as.zoo(table(exception_requests$FormEffectiveDt))

# Calculate the rolling weekly average
exceptions_rolling <- rollapplyr(exceptions_ts, width = 7, FUN = mean, fill = NA)

# Convert the rolling average back to a data frame
exceptions_df <- data.frame(date = index(exceptions_rolling) + as.Date("2018-10-18"), 
                            exceptions = as.numeric(exceptions_rolling))

# Create time-trend graph with rolling weekly average
ggplot(data = exceptions_df, aes(x = date, y = exceptions)) +
  geom_smooth(se = FALSE, span = 0.1) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(title = "Rolling Weekly Average of Heart Transplant Exceptions Used",
       x = "Year",
       y = "Exceptions") +
  theme_bw()

ggsave("exception_time_trend.png", width = 7, height = 5)

# Table of exception utilization by year using JustFormHR
```

# Check number of Status 2 exception candidates with IABPs who did not meet hemodynamic requirements
``` {r stat2_hemo}
df <- full_list
df_stat2ex <- df %>% filter(can_stat_active == 2120 & rec_exception_ever == 1)

load("C:/Users/dyjohnson/Box/Heart Data Pipeline/RData files/full_list_pubsaf2206.RData")
JustFormHR <- read_sas("C:/Users/dyjohnson/Box/SAF 2022 Q2/SupplementalData2206/ThoracicRegistration/JustFormHR.sas7bdat")

full_list_stat2ex <- full_list %>%
  filter(PX_ID %in% df_stat2ex$PX_ID) %>%
  left_join(
    JustFormHR %>%
      select(JustId, FormEffectiveDt, FormExpirationDt, ApplicationStatus),
    by = "JustId"
  ) %>%
  select(
    PX_ID,
    unique_date,
    CAN_IABP,
    CANHX_STAT_CD,
    mean_arterial_pressure,
    cardiac_index,
    PCWP,
    SVO2,
    JustId,
    FormEffectiveDt,
    FormExpirationDt,
    ApplicationStatus,
    RequestedCandStatCd,
    Exception
  )

## Add a copy of CANHX_STAT_CD where a candidate's most recent status is carried through temporarily inactive periods
full_list_stat2ex <- full_list_stat2ex %>%
  mutate(can_stat_active = factor(case_when(
    !(CANHX_STAT_CD %in% c(1999, 2999)) ~ CANHX_STAT_CD
    ))) %>%
  group_by(PX_ID) %>%
  fill(can_stat_active, .direction = "down") %>%
  ungroup()

# Remove justification forms that aren't for approved exceptions
full_list_stat2ex <- full_list_stat2ex %>%
  mutate(
    across(
      c(JustId, FormEffectiveDt, FormExpirationDt, RequestedCandStatCd, ApplicationStatus, Exception),
      ~ case_when(Exception == 1 & ApplicationStatus %in% c(8, 10, 12, 13, 14, 16, 19, 27) ~ .)
    )
  )

# Fill application data from JustFormHR through exception period
full_list_stat2ex <- full_list_stat2ex %>% 
  group_by(PX_ID) %>%
  fill(JustId, FormEffectiveDt, FormExpirationDt, RequestedCandStatCd, Exception, ApplicationStatus, .direction = "down") %>%
  mutate(
    across(
      c(JustId, FormEffectiveDt, FormExpirationDt, RequestedCandStatCd, Exception, ApplicationStatus),
      ~ case_when(unique_date <= FormExpirationDt ~ .)
    )
  ) %>%
  ungroup()

# status_approved represents a candidate's approved status and fills forward
full_list_stat2ex <- full_list_stat2ex %>%
  mutate(status_approved = case_when(
    Exception == 1 &
      ApplicationStatus %in% c(8, 10, 12, 13, 14, 16, 19, 27) ~ RequestedCandStatCd
    )
  ) %>%
  group_by(PX_ID) %>%
  fill(status_approved, .direction = "down") %>%
  ungroup()

full_list_stat2ex <- full_list_stat2ex %>%
  mutate(
    rec_exception = if_else(CANHX_STAT_CD == status_approved & unique_date <= FormExpirationDt & unique_date + days(7) >= FormEffectiveDt, 1, 0, missing = 0)
  )

# Add date when candidates were approved for Status 2 exceptions
full_list_stat2ex <- full_list_stat2ex %>%
  mutate(
    stat2ex_date = case_when(
      status_approved == 2120 &
        rec_exception == 1 ~ FormEffectiveDt
    )
  ) %>%
  group_by(PX_ID) %>%
  fill(stat2ex_date, .direction = "up") %>%
  ungroup()

# Filter data to observations <=2 days before being approved for Status 2 exception
full_list_stat2ex <- full_list_stat2ex %>%
  group_by(PX_ID) %>%
  mutate(days_to_stat2ex = as.numeric(difftime(stat2ex_date, unique_date, units = "days"))) %>%
  filter(between(days_to_stat2ex, 0, 2)) %>%
  ungroup() %>%
  group_by(PX_ID, stat2ex_date) %>%
  filter(unique_date == max(unique_date)) %>%
  ungroup()

# Check hemodynamic requirements
full_list_stat2ex <- full_list_stat2ex %>%
  mutate(
    hemo_req = if_else(
      mean_arterial_pressure < 60 |
        cardiac_index < 2.0 |
        PCWP > 15 |
        SVO2 < 50.0,
      1,
      0
    ),
    hemo_req = if_else(is.na(hemo_req), 0, hemo_req)
  )

# Calculate proportion of Status 2 exceptions for candidates who had IABPs and didn't meet the hemodynamic requirements
sum(full_list_stat2ex$CAN_IABP == 1 & full_list_stat2ex$hemo_req == 0)/n_distinct(full_list_stat2ex)
```
